SIRV-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies

### 2024-02-06 paper_draft/main.tex
beats modelled
changes in the paper

\documentclass[tikz,fleqn,12pt]{wlscirep}
\usepackage[english]{babel}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bm}
\usepackage{tikz}
\usepackage{hyperref}
\usepackage[nobiblatex]{xurl}
\usepackage{amsmath}
\DeclareMathAlphabet{\altmathcal}{OMS}{cmsy}{m}{n}
\usepackage{mathptmx}
\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}
\usetikzlibrary{arrows.meta,positioning,fit}
\tikzset{%
   neuron missing/.style={
    draw=none, 
    scale=2,
    text height=0.333cm,
    execute at begin node=\color{black}$\vdots$
  },
}
\urlstyle{same}

@@

\affil[1]{BML Munjal University}
\affil[*]{e-mail: maeghaljain@gmail.com}
\begin{abstract}
The outbreak of COVID-19 has highlighted the intricate interplay between public health and economic stability on a global scale. This paper aims to develop a reinforcement learning framework to balance health and economic outcomes during infectious disease outbreaks. The framework utilizes the SIR model without vital dynamics and incorporates globally comparable government responses. The study acknowledges the limitations of deterministic models and proposes the use of deep reinforcement learning to reduce input dimensionality and normalize input. While the model offers transparency in terms of its dependency on the reward policy defined, it recognizes the need for a comprehensive consideration of decision factors beyond pure reinforcement learning results.
\end{abstract}
\begin{document}
\flushbottom
\maketitle
\thispagestyle{empty}
\subsection*{Introduction}
In the past, global spread of infectious diseases was largely due to colonization, slavery, and war, leading to widespread illness and death from diseases like tuberculosis, polio, smallpox, and diphtheria. Medical advancements, better access to health care, and improved sanitation have worked towards improving the situation of mortality and morbidity linked to infectious diseases in the past twenty years. However, in low and lower-middle income countries the burden of infectious diseases still persists. The rapid pace of urbanization in low and middle-income countries, along with the rise in populations living in crowded, poor-quality homes, has led to new conditions that favor the emergence of infectious diseases~\cite{Baker2022, inequitableworld}.

Recently, the COVID-19 pandemic caused a havoc worlwide. Till date there have been 772 million cases and more than 6 million deaths~\cite{WHOCovid}. The pandemic triggered the sharpest economic recession in modern history with a 3\% decline, much worse than during the 2008-09 financial crisis~\cite{IMFCovid}. As nations grappled with the immediate health crisis, the economic fallout disproportionately affected vulnerable populations and exacerbated existing inequalities. Lockdowns and restrictions imposed to curb the spread of the virus led to widespread unemployment, business closures, and disruptions in global supply chains~\cite{Nicola2020}. The challenges faced by low and lower-middle income countries were particularly acute, highlighting the intricate interplay between public health and economic stability on a global scale~\cite{Gagnon2023}.

The need for a nuanced understanding of how interventions impact both health outcomes and economic indicators became increasingly evident, prompting a comprehensive examination by epidemiologists to assist policy makers~\cite{Anderson2020}. The outbreak of COVID-19 has prompted epidemiologists to research on various aspects, including mobility control~\cite{Song2022, Chinazzi2020}, vaccination strategies~\cite{Nguyen2021, Kim2022}, stringency measures/non-pharmaceutical interventions (NPIs)~\cite{Jalloh2022, Caldwell2021, Ferguson2020}, and financial considerations~\cite{DeFoo2023}. Despite the numerous studies conducted, very few explore how common interventions meet multiple policy objectives or how a precise articulation of the main policy goals directs the selection of the most effective interventions in terms of health and economic results~\cite{Hollingsworth2011, Song2022, Pangallo2023, Ash2022, Ohi2020, PADMANABHAN2021102676}. The economic impact of the COVID-19 pandemic varied between rich and poor countries. Although COVID-19 deaths had a slightly larger negative effect on the Gross Domestic Product (GDP) in advanced economies, this difference was not statistically significant. However, lockdown restrictions were found to have a more damaging impact on economic activity in emerging and developing economies~\cite{Gagnon2023, Redlin2022,Liang2021}. It's also suggested that an increase in COVID-19 cases was associated with the introduction of harsher NPIs and lockdown measures could be relaxed once vaccination rates increase~\cite{Redlin2022, Patel2021}.

Many economists have studied the effect of COVID-19 on the economy of nations~\cite{Gagnon2023, GagnonKorea, Deb2020,Eichenbaum2021}. In advanced economies like Korea, where the stringency index was below the median the recession was milder than other advanced economies like the United Kingdom where the stringency was much higher~\cite{GagnonKorea}, they achieved it mostly with very aggressive testing, contact tracing, and enforced quarantines and isolations~\cite{Lim2023, KoreaMinister}. In India, social distancing and containment measures have been effective in reducing the number of COVID-19 cases but have come with economic costs. Social distancing had the most adverse effect on the economy in areas with high urbanization~\cite{Deb2020}.

In this paper we want to optimize the government policies regarding stringency. Therefore, alongside epidemiological data, we use the measures of globally comparable government responses~\cite{Hale2021}. We use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the time scale is short enough so that can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we can effectively address the real-world conditions and propose a solution that is both effective and extendable. The current model accounts for the recovery reached through vaccination~\cite{Marinov2022, MaurciodeCarvalho2023, Thater, Turkyilmazoglu2022, YALADANDA2022101052}. However, the study has limitations; the deterministic SIR model fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}.

In order to capture competing costs within the environment and achieve a balance between health and economic outcomes, we intend to employ reinforcement learning~\cite{Nguyen2022, Bastani2021, Song2022, Ohi2020, PADMANABHAN2021102676}. Not only does the formulation of the model deal better with competing costs, but it also offers a more transparency behind the reasoning of the decisions being made in such circumstances. When we conceptualize our problem as a reinforcement learning task, an agent is tasked with making decisions in an environment with the aim of optimizing cumulative rewards. This approach relies on trial and error, in contrast to dynamic programming, which assumes complete knowledge of the environment~\cite{Lavet2018}. We make the use of deep reinforcement learning, which is an advancement to reinforcement learning as it allows to reduce the input dimensionality and normalize the input~\cite{Lavet2018, Arulkumaran2017, Henderson_Islam_Bachman_Pineau_Precup_Meger_2018, Mnih2015}. While our model is more transparent in terms of its dependency on the reward policy defined, it has its limitations. A universal optimal policy may not suit diverse socioeconomic contexts due to variations in healthcare resources and economic vulnerabilities across countries, regions, or cities and a comprehensive consideration of decision factors, extending beyond pure reinforcement learning results is needed~\cite{Song2022, Dunn2017, Demir2006}.

\subsection*{Mathematical Formulation}
In SIR-RL, we use a compartmental model to model infectious disease. We iteratively develop this model to fit the actual data better. In an SIR model people of the population are divided based on whether they are yet to come into contact with an infected person (Susceptible), are infectious themselves (Infectious), or have recovered from the infection (Recovered). These compartments create the SIR model which can be represented as follows:

\begin{figure}[H]
\centering    
\begin{tikzpicture}[node distance=1cm, auto,
    >=Latex, 
    every node/.append style={align=center},
    int/.style={draw, minimum size=1cm}]
   \node [int] (S)             {$S$};
   \node [int, right=of S] (I) {$I$};
   \node [int, right=of I] (R) {$R$};
   \coordinate[right=of I] (out);
   \path[->] (S) edge node {$\lambda$} (I)
             (I) edge node {$\gamma$} (out);
\end{tikzpicture}
\end{figure}
\begin{equation}
  \frac{d S}{d t}=-\lambda S
  \label{eq:S_without_lockdown}
\end{equation}
\begin{equation}
  \frac{d I}{d t}=\lambda S-\gamma I
  \label{eq:I_without_lockdown}
\end{equation}
\begin{equation}
  \frac{d R}{d t}=\gamma I
  \label{eq:R_without_lockdown}
\end{equation}

@@

At the peak of an epidemic $R_0$ declines as there are no more susceptible people left in the pool, therefore, $R_{eff}$ (effective reproductive number) comes into play. $R_{eff}$ is defined as the average number of secondary cases arising from an infected case, at a given point in an epidemic - therefore, it takes into account the existing immunity of the system.

\begin{equation}
  R_{eff} = R_0 \frac{S}{N}
  \label{eq:R_effective}
\end{equation}

$S$ is the number of susceptible people, $N$ is the total population. At the start of an epidemic when everyone is susceptible, $R_{eff} = R_0$ as, $S = N$ (i.e. the whole population is susceptible). $\beta$ and $\gamma$ are also used to define probability of and infectious individual infecting another individual $\beta / (\beta + \gamma)$ and the probability of recovery, $\gamma / (\beta + \gamma)$.

Most government policies look at the value of $R_{eff}$ to come us with an effective strategy to combat the disease as the fate of the evolution of the disease depends upon it. When $R_{eff}$ is less than one, the infected population $I$ will steadily decline to zero. Conversely, if $R_{eff}$ is greater than one, the infected population will increase. In other words, when $\frac{dI(t)}{dt} < 0 \Rightarrow R_{eff} < 1$ and $\frac{dI(t)}{dt} > 0 \Rightarrow R_{eff} > 1},$ the effective reproductive rate $R_{eff}$ serves as a critical threshold that determines whether an infectious disease will rapidly extinguish or escalate into an epidemic~\cite{Cooper2020}.

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from May, 2020, to October, 2022, we simply define the cost to calibrate the model using the huber loss~\cite{huberloss}:

\begin{equation}
  L_{\delta}(y, f(x)) = 
  \begin{cases}
    \frac{1}{2}{(y - f(x))^2} & \text{for } |y - f(x)| \leq \delta, \\
    \delta \cdot (|y - f(x)| - \frac{1}{2}\delta) & \text{otherwise.}
  \label{eq:huberloss}
  \end{cases}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R})
  \label{eq:cost_SIR_model}
\end{equation}

Where, $S$ is the number of susceptible people and $\hat{S}$ is the predicted number of susceptible people, similarly, $I$ for infected and $\hat{I}$ for the predicted number of infected people, and $R$ for recovered. Minimizing this cost function using the Nelder-Mead method\cite{Gao2010}, we get:

\begin{equation}
  \beta_{optimal} = 0.04208479828695971
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.02388356686032017
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.7620817917644795
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 85051490.53250012
  \label{eq:cost_without_lockdown}
\end{equation}
See figure \ref{fig:SIR_model_IND} to see how the model compares with the real data.

Now that, we have a model -- we want to say $\beta$ is a time-varying parameter controlled by the stringency index, $s$~\cite{SIRLockdown,SIRLockdown2,analytical_covid_lockdown_model}. The stringency index is a composite measure based on nine response indicators including school closures, workplace closures, and travel bans, rescaled to a value from 0 to 100 (100 = strictest)~\cite{owidcoronavirus}. This index simply records the strictness of government policies and does not measure or imply the appropriateness or effectiveness of a country's response i.e. a higher score does not necessarily mean that a country's response is "better" than others lower on the index.

To define the new time-varying beta that is dependent on the current stringency index, we say,

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - (s(t)/100)) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}

Where, $s(t)$ is the stringency index at time $t$ and is scaled down by a factor of 100 to normalize it and bring it in the range $[0, 1]$. Optimizing this using the Nelder-Mead method we get:

\begin{equation}
  \beta_{optimal} = 0.4013340889432941
  \label{eq:beta_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.09017476605499258
  \label{eq:gamma_optimal_with_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 4.450625230328213
  \label{eq:r0_with_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 98438821.45587364
  \label{eq:cost_with_lockdown}
\end{equation}

Finally, an additional flow from the susceptible population to the can be shown by adding a vaccination rate $\nu$ in the model.

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - (s(t)/100)) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu S
  \label{eq:R_with_lockdown_and_nu}
\end{equation}

Optimizing these equations with the Nelder-Mead method we get:

\begin{equation}
  \beta_{optimal} = 0.40897034072952304
  \label{eq:beta_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.09196829370123338
  \label{eq:gamma_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \nu_{optimal} = 2.9044029843851936e-05
  \label{eq:nu_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 4.446862329077204
  \label{eq:r0_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 94636860.38436058
  \label{eq:cost_with_lockdown_and_nu}
\end{equation}

However, as observed by the value of $\nu_{optimal}$ from \ref{eq:nu_optimal_with_lockdown_and_nu} which is almost negligible (close to zero) and the overestimation of infected individuals in \ref{fig:SIRV_model_with_lockdown_infections_IND} suggests that $\nu$ might be varying with time. This suggests a time-varying vaccination rate, wherein the transition from susceptibility to direct recovery fluctuates with time. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \ref{eq:beta_optimal_with_lockdown_and_nu} and \ref{eq:gamma_optimal_with_lockdown_and_nu}, we optimize the value of $\nu$ given a sub-interval of $[{start}, {start} + {window}]$ where the ${window} = 20$ \cite{Marinov2022}. Here, ${window}$ is a hyperparameter that can be changed so as to reduce the cost function.

>>>>

% to compile: https://tex.stackexchange.com/questions/140845/how-can-i-ignore-latex-error-while-compiling
% xelatex -interaction=nonstopmode main.tex
\documentclass[tikz,fleqn,12pt]{wlscirep}
\usepackage[english]{babel}
\usepackage[autostyle, english = american]{csquotes}
\usepackage[nameinlink]{cleveref}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bm}
\usepackage{tikz}
\usepackage{hyperref}
\usepackage[nobiblatex]{xurl}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{mathptmx}
\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{nameref}
\DeclareMathAlphabet{\altmathcal}{OMS}{cmsy}{m}{n}
\newcommand{\overbar}[1]{\mkern 1.5mu\overline{\mkern-1.5mu#1\mkern-1.5mu}\mkern 1.5mu}
\usetikzlibrary{arrows.meta,positioning,fit}
% https://texblog.org/2013/05/06/cleveref-a-clever-way-to-reference-in-latex/
% https://tex.stackexchange.com/questions/433971/redefining-ref-such-that-it-is-eqref-when-you-use-refeq-and-then-it-is
% https://tex.stackexchange.com/questions/18986/is-there-a-standard-way-to-refer-to-a-range-of-equations
% \crefname{equation}{}{}
% \creflabelformat{equation}{\textup{(#2#1#3)}}
\newcommand{\crefrangeconjunction}{--}
\MakeOuterQuote{"}

\tikzset{%
   neuron missing/.style={
    draw=none, 
    scale=2,
    text height=0.333cm,
    execute at begin node=\color{black}$\vdots$
  },
}
\urlstyle{same}

@@

\affil[1]{BML Munjal University}
\affil[*]{e-mail: maeghaljain@gmail.com}
\begin{abstract}
The outbreak of COVID-19 has highlighted the intricate interplay between public health and economic stability on a global scale. This study proposes a novel reinforcement learning framework designed to optimize health and economic outcomes during pandemics. The framework leverages the SIR model, integrating both lockdown measures (via a stringency index) and vaccination strategies to simulate disease dynamics. The stringency index, indicative of the severity of lockdown measures, influences both the spread of the disease and the economic health of a country. Developing nations, which bear a disproportionate economic burden under stringent lockdowns, are the primary focus of our study. By implementing reinforcement learning, we aim to optimize governmental responses and strike a balance between the competing costs associated with public health and economic stability. This approach also enhances transparency in governmental decision-making by establishing a well-defined reward function for the reinforcement learning agent. In essence, this study introduces an innovative and ethical strategy to navigate the intricate challenge of balancing public health and economic stability amidst infectious disease outbreaks.
\end{abstract}
\begin{document}
\flushbottom
\maketitle
\thispagestyle{empty}
\section{Introduction}
In the past, global spread of infectious diseases was largely due to colonization, slavery, and war, leading to widespread illness and death from diseases like tuberculosis, polio, smallpox, and diphtheria. Medical advancements, better access to health care, and improved sanitation have worked towards improving the situation of mortality and morbidity linked to infectious diseases in the past twenty years. However, in low and lower-middle income countries the burden of infectious diseases still persists. The rapid pace of urbanization in low and middle-income countries, along with the rise in populations living in crowded, poor-quality homes, has led to new conditions that favor the emergence of infectious diseases~\cite{Baker2022,inequitableworld}.

Recently, the COVID-19 pandemic caused a havoc worldwide. Till date there have been 772 million cases and more than 6 million deaths~\cite{WHOCovid}. The pandemic triggered the sharpest economic recession in modern history with a 3\% decline, much worse than during the 2008-09 financial crisis~\cite{IMFCovid}. As nations grappled with the immediate health crisis, the economic fallout disproportionately affected vulnerable populations and exacerbated existing inequalities. Lockdowns and restrictions imposed to curb the spread of the virus led to widespread unemployment, business closures, and disruptions in global supply chains~\cite{Nicola2020}. The challenges faced by low and lower-middle income countries were particularly acute, highlighting the intricate interplay between public health and economic stability on a global scale~\cite{Gagnon2023}.

The need for a nuanced understanding of how interventions impact both health outcomes and economic indicators became increasingly evident, prompting a comprehensive examination by epidemiologists to assist policy makers~\cite{Anderson2020}. The outbreak of COVID-19 has prompted epidemiologists to research on various aspects, including mobility control~\cite{Song2022,Chinazzi2020}, vaccination strategies~\cite{Nguyen2021, Kim2022}, stringency measures/non-pharmaceutical interventions (NPIs)~\cite{Jalloh2022,Caldwell2021,Ferguson2020}, and financial considerations~\cite{DeFoo2023}. Despite the numerous studies conducted, very few explore how common interventions meet multiple policy objectives or how a precise articulation of the main policy goals directs the selection of the most effective interventions in terms of health and economic results~\cite{Hollingsworth2011, Song2022,Pangallo2023,Ash2022,Ohi2020, PADMANABHAN2021102676,NBERw26981,analytical_covid_lockdown_model}. The economic impact of the COVID-19 pandemic varied between rich and poor countries. Although COVID-19 deaths had a slightly larger negative effect on the Gross Domestic Product (GDP) in advanced economies, this difference was not statistically significant. However, lockdown restrictions were found to have a more damaging impact on economic activity in emerging and developing economies~\cite{Gagnon2023,Redlin2022,Liang2021}. It's also suggested that an increase in COVID-19 cases was associated with the introduction of harsher NPIs and lockdown measures could be relaxed once vaccination rates increase~\cite{Redlin2022,Patel2021}.

Many economists have studied the effect of COVID-19 on the economy of nations~\cite{Gagnon2023,GagnonKorea,Deb2020,Eichenbaum2021}. In advanced economies like Korea, where the stringency index was below the median the recession was milder than other advanced economies like the United Kingdom where the stringency was much higher~\cite{GagnonKorea}, they achieved it mostly with very aggressive testing, contact tracing, and enforced quarantines and isolations~\cite{Lim2023,KoreaMinister}. In India, social distancing and containment measures have been effective in reducing the number of COVID-19 cases but have come with economic costs. Social distancing had the most adverse effect on the economy in areas with high urbanization~\cite{Deb2020}.

In this paper we optimize the government policies regarding stringency. Therefore, alongside epidemiological data, we use the measures of globally comparable government responses~\cite{Hale2021}. We use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the timescale is small enough that it can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we can effectively address the real-world conditions and propose a solution that is both effective and extendable. The current model (SIRV with lockdown) accounts for the recovery reached through vaccination~\cite{Marinov2022,MaurciodeCarvalho2023,Thater,Turkyilmazoglu2022,YALADANDA2022101052} and the effects of lockdown~\cite{SIRLockdown,NBERw26981,NBERw26867}. Traditional SIR model neglects the time-varying property of parameters like $\beta$ and $\gamma$, but in this paper since, lockdown is a parameter of the proposed model, and it changes with time therefore, we propose a time-dependent SIR model\cite{TimeDependentSIR}. However, the study has limitations; First, the deterministic SIR model (predecessor to our proposed model) fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}. Second, the underreporting of cases during the period selected by our study. Lastly, the reinforcement learning agent should be resistant to how the vaccination rate changes -- or to when it is first introduced because then it scopes the environment for succumbing to wishful thinking which can be  potentially dangerous.

After modelling the disease with lockdown (via stringency index) and vaccination, we try to understand the effects of lockdown on the GDP. Therefore, decisions made by the government regarding the level of lockdown to be enforced plays a role on both the public health outcomes and economic stability during a pandemic. On one hand, stringent lockdown measures can effectively slow the spread of the disease, thereby improving public health outcomes. However, these measures often come at the cost of significant economic disruption, leading to job losses, business closures, and reduced economic growth. On the other hand, relaxing lockdown measures may help to mitigate the economic impact of the pandemic, but could result in increased disease transmission and worsened public health outcomes. In order to capture competing costs within the environment and achieve a balance between health and economic outcomes, we intend to employ reinforcement learning~\cite{Nguyen2022,Bastani2021,Song2022,Ohi2020,PADMANABHAN2021102676}. Not only does the formulation of the model deal better with competing costs, but it also offers more transparency behind the reasoning of the decisions being made in such circumstances. When we conceptualize our problem as a reinforcement learning task, an agent is tasked with making decisions in an environment with the aim of optimizing cumulative rewards (i.e., the total amount of reward it receives over the long run). Simply put, given discrete time steps $t = 0, 1, 2, 3, \dots$, at each time step the agent receives a representation of the environment's state, $s_t \in \altmathcal{S}$, and selects an action $a_t \in \altmathcal{A}(s_t)$, where $\altmathcal{A}(s_t)$ is the set of actions available in state $s_t$, and one step later receives a reward $r_{t+1} \in \altmathcal{R}$ and the state is updated~\cite{sutton2018reinforcement}. The way we define the way these rewards that are given to the agent, makes this decision process more transparent, however, it has its limitations. A universal optimal policy may not suit diverse socioeconomic contexts due to variations in healthcare resources and economic vulnerabilities across countries, regions, or cities and a comprehensive consideration of decision factors, extending beyond pure reinforcement learning results is needed~\cite{Song2022,Dunn2017,Demir2006}.

Additionally, since most modern reinforcement learning achievements are due to a combination of deep learning~\cite{Mnih2015}, in the following framework we make the use of this. Deep reinforcement learning is an advancement to reinforcement learning which helps normalize the input and reduce its dimensionality~\cite{Lavet2018,Arulkumaran2017,Henderson_Islam_Bachman_Pineau_Precup_Meger_2018,Mnih2015}. We use a Long Short Term Memory recurrent neural network for time-series data~\cite{NIPS2001_a38b1617,LSTM} and a simple fully connected network for the non-time-series data.

In summary, by using reinforcement learning augmented with deep learning techniques for the SIRV with lockdown environment, we can better understand the effects of lockdown measures on both public health outcomes and economic stability during a pandemic. However, it is crucial to consider the limitations of this approach and take into account a comprehensive set of decision factors in order to make informed policy decisions that are tailored to specific socioeconomic contexts.

\section{Mathematical Formulation and Numerical Computation}
In SIRV-RL, we use a compartmental model to model infectious disease environment. We iteratively develop this model to fit the actual data better. In an SIR model people of the population are divided based on whether they are yet to come into contact with an infected person (Susceptible), are infectious themselves (Infectious), or have recovered from the infection (Recovered). These compartments create the SIR model which can be represented as follows:

\subsection{Simple SIR Model}
\begin{figure}[H]
\centering    
\begin{tikzpicture}[node distance=1cm, auto,
    >=Latex, 
    every node/.append style={align=center},
    int/.style={draw, minimum size=1cm}]
   \node [int] (S)             {$S$};
   \node [int, right=of S] (I) {$I$};
   \node [int, right=of I] (R) {$R$};
   \coordinate[right=of I] (out);
   \path[->] (S) edge node {$\lambda$} (I)
             (I) edge node {$\gamma$} (out);
\end{tikzpicture}
\end{figure}
\begin{equation}
  \frac{d S}{d t}=-\lambda S
  \label{eq:S_without_lockdown}
\end{equation}
\begin{equation}
  \frac{d I}{d t}=\lambda S-\gamma I
  \label{eq:I_without_lockdown}
\end{equation}
\begin{equation}
  \frac{d R}{d t}=\gamma I
  \label{eq:R_without_lockdown}
\end{equation}

@@

At the peak of an epidemic $R_0$ declines as there are no more susceptible people left in the pool, therefore, $R_e$ (effective reproductive number) comes into play. $R_e$ is defined as the average number of secondary cases arising from an infected case, at a given point in an epidemic, therefore, it takes into account the existing immunity of the system.

\begin{equation}
  R_e = R_0 \frac{S}{N}
  \label{eq:R_effective}
\end{equation}

$S$ is the number of susceptible people, $N$ is the total population. At the start of an epidemic when everyone is susceptible, $R_e = R_0$ as, $S = N$ (i.e., the whole population is susceptible). $\beta$ and $\gamma$ are also used to define probability of and infectious individual infecting another individual $\beta / (\beta + \gamma)$ and the probability of recovery, $\gamma / (\beta + \gamma)$.

Most government policies look at the value of $R_e$ to come us with an effective strategy to combat the disease as the fate of the evolution of the disease depends upon it. When $R_e$ is less than one, the infected population $I$ will steadily decline to zero. Conversely, if $R_e$ is greater than one, the infected population will increase. In other words, when $\frac{dI(t)}{dt} < 0 \Rightarrow R_e < 1$ and $\frac{dI(t)}{dt} > 0 \Rightarrow R_e > 1$, therefore, the effective reproductive rate $R_e$ serves as a critical threshold that determines whether an infectious disease will rapidly extinguish or escalate into an epidemic~\cite{Cooper2020}.

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from May, 2020, to October, 2022, we simply define two cost functions \cref{eq:cost_SIR,eq:cost_I} to calibrate the model with the use of huber loss~\cite{huberloss}. We generally calibrate the model using \cref{eq:cost_SIR}. However, sometimes we must strike a balance between modeling all three population groups (susceptible, infected, and recovered) and focusing just on the infected group, which is most important because it drives the spread of the disease, therefore, in that case we consider the loss obtained from \cref{eq:cost_SIR,eq:cost_I} (see \cref{optimizing_window_length}).

\begin{equation}
  L_{\delta}(y, f(x)) = 
  \begin{cases}
    \frac{1}{2}{(y - f(x))^2} & \text{for } |y - f(x)| \leq \delta, \\
    \delta \cdot (|y - f(x)| - \frac{1}{2}\delta) & \text{otherwise.} \\
  \end{cases}
  \label{eq:huberloss}
\end{equation}

In the above equation, $y$ is the actual data and $f(x)$ is the prediction.

\begin{equation}
  \begin{split}
  \textrm{loss\_SIR} = \textrm{cost\_function\_SIR}(S, \hat{S}, I, \hat{I}, R, \hat{R}) \\
  = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R})
  \end{split}
  \label{eq:cost_SIR}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = \textrm{cost\_function\_I}(I, \hat{I}) = L_{\delta = 1}(I, \hat{I})
  \label{eq:cost_I}
\end{equation}

Where, $S$ is the number of susceptible people and $\hat{S}$ is the predicted number of susceptible people, similarly, $I$ for infected and $\hat{I}$ for the predicted number of infected people, and $R$ for recovered. Using the equations \Crefrange{eq:S_without_lockdown}{eq:R_without_lockdown} , we minimize the cost function \cref{eq:cost_SIR_without_lockdown} using the Nelder-Mead method\cite{Gao2010} to estimate the parameters like $\beta$, $\gamma$, and fit the model to actual data. The following parameters and loss is obtained:

\begin{equation}
  \beta_{optimal} = 0.042
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.024
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.762
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 85051490.533
  \label{eq:cost_SIR_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 45187665.281
  \label{eq:cost_I_without_lockdown}
\end{equation}

See \cref{fig:SIR_model_IND} to see how the model compares with the actual data.

\subsection{SIR Model with Lockdown}
Now that, a simple SIR model has been established -- we need to model the effects of the stringency index (measure for the strictness of lockdown) on $\beta$ (the effective transmission rate). To do this, we say the flow of susceptibles not only depend on $\beta$ but also $s(t)$ the stringency index at time~\cite{NBERw26867,AlvarezLockdownSIR,SIRLockdown,SIRLockdown2,analytical_covid_lockdown_model}. The stringency index is a composite measure based on nine response indicators including school closures, workplace closures, and travel bans, rescaled to a value from 0 to 100 (100 = strictest)~\cite{owidcoronavirus}. This index simply records the strictness of government policies and does not measure or imply the appropriateness or effectiveness of a country's response i.e. a higher score does not necessarily mean that a country's response is "better" than others lower on the index.

To define the new time-varying beta that is dependent on the current stringency index, the following equations have been formulated:

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - s(t)/100) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}

Where, $s(t)$ is the stringency index at time $t$ and is scaled down by a factor of 100 to normalize it and bring it in the range $s(t)/100 \in [0, 1]$. Multiplying the rate of flow from $S$ to $I$ compartment with $1 - s(t)/100$ allows us to account for the effect that stringency has on the disease progression. A higher stringency index can theoretically, stop the flow from the susceptible population to the infected population entirely. Optimizing these equations with \cref{eq:cost_SIR} using the Nelder-Mead method, we get the following parameters and loss:

\begin{equation}
  \beta_{optimal} = 0.401
  \label{eq:beta_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.090
  \label{eq:gamma_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \begin{split}
    \overbar{R_0} = 1.693 \quad \text{(Mean)} \\
    \widetilde{R_0} = 1.624 \quad \text{(Median)} \\
    \text{Mode}(R_0) = 0.804 \quad \text{(Mode)} \\
    \sigma_{R_0} = 0.786 \quad \text{(Standard Deviation)} \\
    R_0 \in [0.16467, 3.0497] \quad \text{(Range)}
  \end{split}
  \label{eq:r0_with_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 98438821.456
  \label{eq:cost_SIR_with_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 11345389.686
  \label{eq:cost_I_with_lockdown}
\end{equation}

\subsection{SIRV Model with Lockdown}
Lastly, an additional flow from the susceptible population to the recovered population can be shown by adding a vaccination rate $\nu$ in the model.

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - s(t)/100) \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu S
  \label{eq:R_with_lockdown_and_nu}
\end{equation}

Optimizing these equations with \cref{eq:cost_SIR} using the Nelder-Mead method:
